FAIR fission track analysis with geochron@home

### 2025-09-28 geochron@home/manuscript/paper.tex
Added dot plot.

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable}, for
the same selection of 25 grains that were analysed by AC and PV in
Section~\ref{sec:private}, along with corresponding counts made by
PV. Everyone counted tracks in the same ROIs. The vast majority of
students counted fewer tracks than PV did, often many fewer, and on
two occasions someone counted no tracks at all. PV's count is always
above the students' median and nearly always above the upper quartile.
Table~\ref{tab:crowdtable} (looking along rows) shows just 7 students
with average counts within 10\% of PV's, and 18 with average counts
less than half of his.\medskip

However, in spite of the under-counting by the students, the
\emph{ratios} of the track counts between different grains are often
close to PV's ratios, especially when comparing grains with similar
track densities.  This is apparent from Table~\ref{tab:crowdtable}
(looking along columns).  In grains containing few tracks (left hand
side of the table), the median of the student counts is within 20\% of
PV's counts.  However, this gradually levels off to 50\% at higher
track densities (right hand side of the
table). Figure~\ref{fig:23vs25}a shows a scatter plot of pairs of
counts for two grains (23 and 25) made by 36 students who counted both
of them. The two grains were chosen because they have similar track
counts for both PV and the median student. They both have a median of
28 tracks per grain compared with PV's pair of 52 and 50, which is
highlighted in blue. There is a positive correlation between the
pairs, consistent with systematic observer effects (i.e., people who
count a low value in one grain tend to count similarly low in the
other, and vice versa).  Comparing other pairs of grains shows similar
results, with correlations varying between 0.4 and 0.9. In all cases
we have looked at, the residual variation, after allowing for
systematic differences between students, is consistent with Poisson
variation.\medskip

@@

\begin{small}
  \input{../output/shortcrowdtable.txt}\\
\end{small}
\captionof{table}{Counts of fission tracks in the same ROIs of 25
  grains made by PV and 68 students. Only 20 grains and 20 students
  are shown due to space constraints; the full table is provided in
  the supplementary information.  Each student counted a subset of the
  grains. The grains are listed in increasing order of PV's
  counts. The last 3 columns give, for each student, the number of
  grains counted $n_g$ and the weighted mean $r_p =
  \sum\,w_i\,r_i/\sum\,w_i$ and standard deviation $s_p =
  \sqrt{w_i\,(r_i-r_p)^2/(n_g − 1)}$ where $r_i$ is the ratio of the
  student's count to PV's count for grain $i$ and $w_i$ is PV's count
  for grain $i$. ($r_p$ is the same as the ratio of the student's
  total count over the $n_g$ grains to PV's total count over the same
  grains.)  The students are listed in decreasing order of $r_p$ . The
  last 3 rows ($n$, $\bar{y}$ and $s_y$ ) give the sample size (number
  of students), mean and standard deviation of the students' counts
  for each grain.}
\label{tab:crowdtable}
%\end{minipage}

>>>>

\begin{small}
  \input{../output/shortcrowdtable.txt}\\
\end{small}
\captionof{table}{Counts of fission tracks in the same ROIs of 25
  grains made by PV and 68 students. Only 20 students are shown due to
  space constraints; the full table is provided in the supplementary
  information.  Each student counted a subset of the grains. The
  grains are listed in increasing order of PV's counts. The last 3
  columns give, for each student, the number of grains counted $n_g$
  and the weighted mean $r_p = \sum\,w_i\,r_i/\sum\,w_i$ and standard
  deviation $s_p = \sqrt{\sum w_i\,(r_i-r_p)^2/(n_g − 1)}$
  where $r_i$ is the ratio of the student's count to PV's count for
  grain $i$ and $w_i$ is PV's count for grain $i$. ($r_p$ is the same
  as the ratio of the student's total count over the $n_g$ grains to
  PV's total count over the same grains.)  The students are listed in
  decreasing order of $r_p$. $r_m$ is the ratio of the students'
  median count per grain to PV's count.}
\label{tab:crowdtable}

The raw counts for 68 students (two of the students were unable to
complete the exercise) are given in Table~\ref{tab:crowdtable} and
Figure~\ref{fig:stripchart}, for the same selection of 25 grains that
were analysed by AC and PV in Section~\ref{sec:private}, along with
corresponding counts made by PV. Everyone counted tracks in the same
ROIs. Variation between the mean values is to be expected because of
the differing numbers of tracks due to varying areas and U contents of
the grains. But the variation between counts within grains (shown by
the standard deviations in Figure~\ref{fig:stripchart}) is due
entirely to differences between the students' recognising and counting
exactly the same tracks. These standard deviations increase with the
mean number of tracks and are considerable in size, being on average
about 40\% of the mean. With expert trained counters one would expect
much smaller differences between counts.  Furthermore, the vast
majority of students counted fewer tracks than PV did, often many
fewer, and on two occasions someone counted no tracks at all. PV's
count is always above the students' median and nearly always above the
upper quartile (Figure~\ref{fig:stripchart}).
Table~\ref{tab:crowdtable} (looking along rows) shows just 7 students
with average counts within 10\% of PV's, and 18 with average counts
less than half of his.\medskip

The rightmost three columns of Table~\ref{tab:crowdtable} contain
summary statistics ($n_g$ , $r_p$ and $s_p$) giving, for each student,
the number of grains counted and the weighted mean and standard
deviation of the ratios of the student's count to PV's count for each
grain (definitions provided in the caption). The true number of tracks
in each ROI is unknown, so we are comparing the students with PV
(remembering that PV's count is also subject to error, which is the
same for each student). $r_p$ measures agreement with PV on average
and $s_p$ measures consistency over grains. For good agreement $r_p$
should be close to 1 and $s_p$ should be small.  Perfect agreement is
$r_p = 1$ and $s_p = 0$. For comparison, an $r_p = 1.04$ and $s_p =
0.69$ for PV and AC using just the counts in the common areas $A_0$
from Section~\ref{sec:private}. With the caveat that the ROIs are
different and PV's counts are different, this calculation shows that
there is a group of 6 (possibly 9) students whose agreement with PV is
comparable with AC's. This group does not include student 229 whose
$r_p = 0.95$ but with a large $s_p = 1.62$.\medskip

However, in spite of the under-counting by the students, the ratio of
their pooled track counts ($r_p$) for two different grains is often
close to PV's ratio, especially when comparing grains with similar
track densities. This is illustrated in Figure~\ref{fig:23vs25}a for
grains~23 and 25.  For any point, the ratio of the two counts is given
by the slope of a line from the origin to that point. Although the
individual slopes vary considerably, the line with slope equal to the
ratio of the total counts for all students (which is 0.96) passes
close to PV's pair of counts (52,50) shown by the blue dot (with ratio
1.03). It is interesting that the students with lower counts are
mostly above this line and those with higher counts are all below it.
There is a positive correlation between the pairs of counts,
consistent with systematic observer effects (i.e., people who count a
low value in one grain tend to count similarly low in the other, and
vice versa). Comparing other pairs of grains shows similar results,
with correlations varying between 0.4 and 0.9.\medskip

\noindent\begin{minipage}{.55\linewidth}
  \includegraphics[width=\linewidth]{stripchart.pdf}\medskip
\end{minipage}
\begin{minipage}{.45\linewidth}
\captionof{figure}{ Dot plot of the data shown in
  Table~\ref{tab:crowdtable}, with the students' medians and PV's
  counts shown as yellow and blue triangles, respectively.
  \medskip}
  \label{fig:stripchart}
\end{minipage}

@@
